Point estimates comparison for the simplified Naomi model

### 2023-07-03 src/naomi-simple_point-estimates/point-estimates.Rmd
Add point estimates notebook (and split out latent and output analysis)

# Means and standard deviations

@@

type_key <- data.frame(
  par = c(latent_pars, output_pars),
  type = c(rep("latent", times = length(latent_pars)), rep("output", times = length(output_pars)))
)

df_tmb <- lapply(tmb$fit$sample[type_key$par], function(row) {
  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
    tibble::rowid_to_column("id")
  }) %>%
  dplyr::bind_rows(.id = "par") %>%
  mutate(method = "TMB")

df_aghq <- lapply(aghq$quad$sample[type_key$par], function(row) {
  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
    tibble::rowid_to_column("id")
  }) %>%
  dplyr::bind_rows(.id = "par") %>%
  mutate(method = "PCA-AGHQ")

df_tmbstan <- lapply(tmbstan$mcmc$sample[type_key$par], function(row) {
  data.frame("sd" = matrixStats::rowSds(row), "mean" = rowMeans(row)) %>%
    tibble::rowid_to_column("id")
  }) %>%
  dplyr::bind_rows(.id = "par") %>%
  mutate(method = "NUTS")

@@

df_plot <- df %>%
  pivot_longer(cols = c("sd", "mean"), names_to = "indicator", values_to = "estimate") %>%
  pivot_wider(values_from = "estimate", names_from = "method") %>%
  pivot_longer(cols = c("TMB", "PCA-AGHQ"), names_to = "method", values_to = "approximate") %>%
  rename("truth" = "NUTS") %>%
  mutate(
    indicator = fct_recode(indicator, "Posterior mean estimate" = "mean", "Posterior SD estimate" = "sd"),
    method = fct_relevel(method, "TMB", "PCA-AGHQ")
  )

df_metrics <- df_plot %>%
  group_by(method, indicator) %>%
  summarise(
    rmse = sqrt(mean((truth - approximate)^2)),
    mae = mean(abs(truth - approximate))
  )

df_metrics_pct <- df_metrics %>%
  ungroup() %>%
  group_by(indicator) %>%
  summarise(
    rmse_diff = 100 * diff(rmse) / max(rmse),
    mae_diff = 100 * diff(mae) / max(mae)
  )

df_metrics <- df_metrics %>%
  mutate(
    label = ifelse(
      method == "PCA-AGHQ",
      paste0("RMSE: ", round(rmse, 2), " (", round(df_metrics_pct$rmse_diff), "%)", "\nMAE: ", round(mae, 2), " (", round(df_metrics_pct$mae_diff), "%)"),
      paste0("RMSE: ", round(rmse, 2), "\nMAE: ", round(mae, 2))
    )
  )

write_csv(df_metrics, "mean-sd.csv")

mean_sd_plot <- ggplot(df_plot, aes(x = truth, y = approximate - truth)) +
  geom_point(shape = 1, alpha = 0.4) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
  geom_text(data = df_metrics, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
  labs(x = "NUTS", y = "Approximation - NUTS") +
  theme_minimal()

ggsave("mean-sd.png", mean_sd_plot, h = 6, w = 6.25)

#' Split into two plots for presentations etc.

jitter_amount <- 0.02

mean_plot <- df_plot %>%
  filter(indicator == "Posterior mean estimate") %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_jitter(shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  lims(y = c(-0.4, 0.4)) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
  geom_text(
    data = filter(df_metrics, indicator == "Posterior mean estimate"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "NUTS", y = "Approximation - NUTS") +
  theme_minimal()

ggsave("mean.png", mean_plot, h = 4, w = 6.25)

sd_plot <- df_plot %>%
  filter(indicator == "Posterior SD estimate") %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_jitter(shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  lims(y = c(-0.6, 0.6)) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
  geom_text(
    data = filter(df_metrics, indicator == "Posterior SD estimate"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "NUTS", y = "Approximation - NUTS") +
  theme_minimal()

ggsave("sd.png", sd_plot, h = 4, w = 6.25)

sd_plot_alt <- sd_plot +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
  )

y_axis <- ggplot(data.frame(l = mean_plot$labels$y, x = 1, y = 1)) +
  geom_text(aes(x, y, label = l), angle = 90) +
  theme_void() +
  coord_cartesian(clip = "off")

mean_plot_alt <- mean_plot
mean_plot_alt$labels$x <- ""
mean_plot_alt$labels$y <- sd_plot_alt$labels$y <- ""

mean_sd_plot_alt <- y_axis + (mean_plot_alt / sd_plot_alt) +
  plot_layout(widths = c(1, 30))

ggsave("mean-sd-alt.png", mean_sd_plot_alt, h = 6, w = 6.25)
```

>>>>

# Data manipulation

We're interested in the latent field and certain model outputs (prevalence, incidence and treatment coverage):

@@

type_key <- data.frame(
  par = c(latent_pars, output_pars),
  type = c(rep("latent", times = length(latent_pars)), rep("output", times = length(output_pars)))
)
```

Extract the mean and standard deviation for each parameter for each inference method, then bind them together:

```{r}
df_tmb <- lapply(tmb$fit$sample[type_key$par], function(row) {
  data.frame("mean" = rowMeans(row), "sd" = matrixStats::rowSds(row)) %>%
    tibble::rowid_to_column("id")
  }) %>%
  dplyr::bind_rows(.id = "par") %>%
  mutate(method = "TMB")

df_aghq <- lapply(aghq$quad$sample[type_key$par], function(row) {
  data.frame("mean" = rowMeans(row), "sd" = matrixStats::rowSds(row)) %>%
    tibble::rowid_to_column("id")
  }) %>%
  dplyr::bind_rows(.id = "par") %>%
  mutate(method = "PCA-AGHQ")

df_tmbstan <- lapply(tmbstan$mcmc$sample[type_key$par], function(row) {
  data.frame("mean" = rowMeans(row), "sd" = matrixStats::rowSds(row)) %>%
    tibble::rowid_to_column("id")
  }) %>%
  dplyr::bind_rows(.id = "par") %>%
  mutate(method = "NUTS")

@@

head(df)
```

Pivot so that the data is ready to plot:

```{r}
df_plot <- df %>%
  pivot_longer(cols = c("sd", "mean"), names_to = "indicator", values_to = "estimate") %>%
  pivot_wider(values_from = "estimate", names_from = "method") %>%
  pivot_longer(cols = c("TMB", "PCA-AGHQ"), names_to = "method", values_to = "approximate") %>%
  rename("truth" = "NUTS") %>%
  mutate(
    indicator = fct_recode(indicator, "Posterior mean estimate" = "mean", "Posterior SD estimate" = "sd"),
    method = fct_relevel(method, "TMB", "PCA-AGHQ")
  )

head(df_plot)
```
Calculate the root mean square and mean absolute errors between the approximate methods and NUTS (taken to be the truth):

```{r}
df_metrics <- df_plot %>%
  group_by(method, indicator, type) %>%
  summarise(
    rmse = sqrt(mean((truth - approximate)^2)),
    mae = mean(abs(truth - approximate))
  ) %>%
  ungroup()

head(df_metrics)
```

Add difference in metric between TMB and PCA-AGHQ:

```{r}
df_metrics_pct <- df_metrics %>%
  split(~ indicator + type) %>%
  lapply(function(x) {
    x %>% summarise(
      rmse_diff = 100 * diff(rmse) / max(rmse),
      mae_diff = 100 * diff(mae) / max(mae)
    )
  }) %>%
  bind_rows(.id = "id") %>%
  separate(id, c("indicator", "type"), sep = "\\.")

#' Add as a naming column for plot
df_metrics <- df_metrics %>%
  left_join(
    df_metrics_pct,
    by = c("indicator", "type")
  ) %>%
  mutate(
    label = ifelse(
      method == "PCA-AGHQ",
      paste0("RMSE: ", signif(rmse, 2), " (", signif(rmse_diff, 2), "%)", "\nMAE: ", signif(mae, 2), " (", signif(mae_diff, 2), "%)"),
      paste0("RMSE: ", signif(rmse, 2), "\nMAE: ", signif(mae, 2))
    )
  )

write_csv(df_metrics, "mean-sd.csv")
```

# Plots

## Latent field

```{r}
df_plot_latent <- filter(df_plot, type == "latent")
df_metrics_latent <- filter(df_metrics, type == "latent")

(mean_sd_plot_latent <- df_plot_latent %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_point(shape = 1, alpha = 0.4) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
  geom_text(data = df_metrics_latent, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
  labs(x = "NUTS", y = "Approximation - NUTS") +
  theme_minimal())

ggsave("mean-sd-latent.png", mean_sd_plot_latent, h = 6, w = 6.25)
```

Version split into two for presentations

```{r}
jitter_amount <- 0.02

mean_plot_latent <- df_plot_latent %>%
  filter(indicator == "Posterior mean estimate") %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_jitter(shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  lims(y = c(-0.4, 0.4)) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
  geom_text(
    data = filter(df_metrics_latent, indicator == "Posterior mean estimate"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "NUTS", y = "Approximation - NUTS") +
  theme_minimal()

ggsave("mean-latent.png", mean_plot_latent, h = 4, w = 6.25)

sd_plot_latent <- df_plot_latent %>%
  filter(indicator == "Posterior SD estimate") %>%
  ggplot(aes(x = truth, y = approximate - truth)) +
  geom_jitter(shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  lims(y = c(-0.6, 0.6)) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
  geom_text(
    data = filter(df_metrics_latent, indicator == "Posterior SD estimate"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "NUTS", y = "Approximation - NUTS") +
  theme_minimal()

ggsave("sd-latent.png", sd_plot_latent, h = 4, w = 6.25)

sd_plot_latent_alt <- sd_plot_latent +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
  )

y_axis <- ggplot(data.frame(l = mean_plot_latent$labels$y, x = 1, y = 1)) +
  geom_text(aes(x, y, label = l), angle = 90) +
  theme_void() +
  coord_cartesian(clip = "off")

mean_plot_latent_alt <- mean_plot_latent
mean_plot_latent_alt$labels$x <- ""
mean_plot_latent_alt$labels$y <- sd_plot_latent_alt$labels$y <- ""

(mean_sd_plot_latent_alt <- y_axis + (mean_plot_latent_alt / sd_plot_latent_alt) +
  plot_layout(widths = c(1, 30)))

ggsave("mean-sd-alt-latent.png", mean_sd_plot_latent_alt, h = 6, w = 6.25)
```

## Outputs

```{r}
df_plot_output <- filter(df_plot, type == "output")
df_metrics_output <- filter(df_metrics, type == "output")
cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

(mean_sd_plot_output <- df_plot_output %>%
  ggplot() +
  geom_point(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4) +
  scale_color_manual(values = cbpalette) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed") +
  geom_text(data = df_metrics_output, aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5) +
  labs(x = "NUTS", y = "Approximation - NUTS") +
  theme_minimal())

ggsave("mean-sd-output.png", mean_sd_plot_output, h = 6, w = 6.25)
```

Version split into two for presentations

```{r}
jitter_amount <- 0.02

mean_plot_output <- df_plot_output %>%
  filter(indicator == "Posterior mean estimate") %>%
  ggplot() +
  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  scale_color_manual(values = cbpalette) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
  geom_text(
    data = filter(df_metrics_output, indicator == "Posterior mean estimate"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "NUTS", y = "Approximation - NUTS", col = "Output") +
  theme_minimal()

ggsave("mean-output.png", mean_plot_output, h = 4, w = 6.25)

sd_plot_output <- df_plot_output %>%
  filter(indicator == "Posterior SD estimate") %>%
  ggplot() +
  geom_jitter(aes(x = truth, y = approximate - truth, color = par), shape = 1, alpha = 0.4, width = jitter_amount, height = jitter_amount) +
  scale_color_manual(values = cbpalette) +
  facet_grid(indicator ~ method) +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed", size = 0.25) +
  geom_text(
    data = filter(df_metrics_output, indicator == "Posterior SD estimate"),
    aes(x = -Inf, y = Inf, label = label), size = 3, hjust = 0, vjust = 1.5
  ) +
  labs(x = "NUTS", y = "Approximation - NUTS", col = "Output") +
  theme_minimal()

ggsave("sd-output.png", sd_plot_output, h = 4, w = 6.25)

sd_plot_output_alt <- sd_plot_output +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    legend.position = "bottom"
  )

y_axis <- ggplot(data.frame(l = mean_plot_output$labels$y, x = 1, y = 1)) +
  geom_text(aes(x, y, label = l), angle = 90) +
  theme_void() +
  coord_cartesian(clip = "off")

mean_plot_output_alt <- mean_plot_output +
  theme(legend.position = "none")

mean_plot_output_alt$labels$x <- ""
mean_plot_output_alt$labels$y <- sd_plot_output_alt$labels$y <- ""

(mean_sd_plot_output_alt <- y_axis + (mean_plot_output_alt / sd_plot_output_alt) +
  plot_layout(widths = c(1, 30)))

ggsave("mean-sd-alt-output.png", mean_sd_plot_output_alt, h = 6, w = 6.25)
```
